---
title: "Mutually exclusive transit service areas"
author: "Carole Voulgaris"
date: "4/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(tidycensus)
library(here)
library(leaflet)
library(htmlwidgets)
library(tidytransit)
```

The goal of this analysis is to generate a set of transit route service areas such that every census block in the study area is assigned to the service area of exactly one transit route.

The study area is the set of census tracts in Los Angeles County with 500 meters of a transit stop.

#Load ridership data

Stop level ridership data for October 2019 are from https://la-metro.maps.arcgis.com/home/item.html?id=a1d134368e5e4fe48087b187da7354d2. This is a shapefile with the total number of boardings and alightings on each route served by each transit stop. 

This analysis requires us to assign each transit stop to exactly one route. Many transit stops (3374 out of 14058, or 24 percent) serve more than one route, in these cases, the stop was assigned to the route with the highest activity (boardings plus alightings) at that stop.

```{r}
ridership <- here("02-data",
                  "LA_metro_oct2019.csv") %>%
  st_read(options=c("X_POSSIBLE_NAMES=long_",
                    "Y_POSSIBLE_NAMES=lat")) %>%
  st_set_crs("WGS84") %>%
   select(stopnum, 
          DX_ACT_1,
          DX_ACT_2,
          DX_ACT_3,
          DX_ACT_4,
          DX_ACT_5,
          DX_ACT_6,
          DX_ACT_7,
          DX_ACT_8,
          DX_ACT_9,
          DX_ACT_10,
          DX_ACT_11,
          DX_ACT_12,
          DX_ACT_13,
          DX_ACT_14,
          DX_ACT_15,
          DX_ACT_16,
          DX_ACT_17,
          DX_ACT_18,
          LINEDIR_1,
          LINEDIR_2,
          LINEDIR_3,
          LINEDIR_4,
          LINEDIR_5,
          LINEDIR_6,
          LINEDIR_7,
          LINEDIR_8,
          LINEDIR_9,
          LINEDIR_10,
          LINEDIR_11,
          LINEDIR_12,
          LINEDIR_13,
          LINEDIR_14,
          LINEDIR_15,
          LINEDIR_16,
          LINEDIR_17,
          LINEDIR_18) 

activity <- ridership %>%
  select(stopnum, starts_with("DX_ACT")) %>%
  pivot_longer(cols = starts_with("DX_ACT"),
               names_to = "line",
               names_prefix = "DX_ACT_",
               values_to = "DX_ACT") %>%
  filter(DX_ACT != "") %>%
  mutate(DX_ACT = as.numeric(DX_ACT)) %>%
  st_drop_geometry()

lines <- ridership %>%
  select(stopnum, starts_with("LINEDIR")) %>%
  pivot_longer(cols = starts_with("LINEDIR"),
               names_to = "line",
               names_prefix = "LINEDIR_",
               values_to = "LINEDIR") %>%
  filter(LINEDIR != "") 

route_locs <- full_join(lines, activity) %>%
  mutate(route = str_split_fixed(LINEDIR, "-", 2)[,1]) %>%
  group_by(stopnum) %>%
  mutate(max_riders = max(DX_ACT)) %>%
  ungroup() %>%
  filter(DX_ACT == max_riders) %>%
  select(route)
  
```

# Define study area

The study area is defined as the set of tracts within 500 meters of an LA Metro transit stop. This represents 41 percent of land area of Los Angeles County and 76 percent of the population.

```{r}
tract_pop <- get_decennial(geography = "tract",
                           state = "CA", 
                           county = "Los Angeles",
                           variables = "P1_001N",
                           year = 2020,
                           output = "wide") 

tracts <- tracts(state = "CA", 
                 county = "Los Angeles") %>%
  st_transform("WGS84") %>%
  left_join(tract_pop, by = "GEOID")

county_pop = sum(tracts$P1_001N)
county_area = sum(tracts$ALAND)

tracts <- tracts %>%
  filter(lengths(st_is_within_distance(tracts, route_locs, dist = 500)) > 0)

tracts %>%
  select(GEOID) %>%
  st_drop_geometry() %>%
  write_csv(file = here("02-data",
                        "study-area-tracts.csv"))

pct_pop_in_area = sum(tracts$P1_001N) / county_pop
pct_county_in_area = sum(tracts$ALAND) / county_area
```

## Map study area

Here is an interactive map of the census tracts in the study area.

```{r}
map <- leaflet(tracts) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = TRUE),
      weight = 1,
      fillColor = "green",
      color = "green",
      label = ~ GEOID)

here("03-maps",
     "tracts.html") %>%  
  saveWidget(widget = map, .)
```

# Load census blocks

```{r}
blocks <- blocks(state = "CA", county = "Los Angeles") %>%
  st_transform("WGS84") %>%
  mutate(tract = substr(GEOID20, 1, 11))

blocks <- blocks %>%
  filter(tract %in% tracts$GEOID) %>%
  st_join(route_locs, join = st_nearest_feature) %>%
  select(GEOID20, route) 
```

# Create route zone boundaries

The summarize function takes a very long time (a little over an hour). I'm saving the result so you can just read it from file and skip this chunk.

```{r, eval=FALSE}
route_zones <- blocks %>%
  group_by(route) %>%
  summarise()

here("02-data",
     "route-zones.geojson") %>%  
  st_write(obj = route_zones, .)
```


## Map the boundaries

```{r}
route_zones <- st_read("route-zones.geojson")

map <- leaflet(route_zones) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = FALSE),
      weight = 1,
      color = "black",
      fillColor = "green",
      label = ~ route) 

here("03-maps",
     "route-zones.html") %>%  
  saveWidget(widget = map, .)
```

# Save the list of blocks

```{r}
blocks %>%
  st_drop_geometry() %>%
  write_csv(here("02-data",
                 "blocks-by-route.csv"))
```

