---
title: "Mutually exclusive transit service areas"
author: "Carole Voulgaris Gabriel Barrett-Jackson"
date: "12/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(tidycensus)
library(leaflet)
library(htmlwidgets)
library(tidytransit)
library(utils)
library(magrittr)
library(here)
```

```{r}
#tot_pop <- "P001001"
blk_pop <- "P003003"
hispanic_pop <- "P005001"
asian_pop <- "P003005"
white_pop <- "P005003" 
nat_am_pop <- "P003004"
nhpi_pop <- "P003006"

demographics <- c(
   #tot_pop = "P001001", 
   blk_pop = "P003003", 
   hispanic_pop = "P005001", 
   asian_pop = "P003005",
   white_pop = "P005003", 
   nat_am_pop = "P003004",
   nhpi_pop = "P003006")
   
#non_white_pop <- c(
   #blk_pop = "P003003", 
   #hispanic_pop = "P004001", 
   #asian_pop = "P003005",
   #nat_am_pop = "P003004",
   #nhpi_pop = "P003006")
```

The goal of this analysis is to generate a set of transit route service areas such that every census block in the study area is assigned to the service area of exactly one transit route.

The study area is the set of zip code tabulation areas in Los Angeles, Chicago, Dallas, and Boston with 500 meters of a transit stop.

Steps to doing this with GTFS:

You want to end up with a data frame with each stop in one row, and a column indicating the route that stops there most frequently.

What you start with:
stops.txt: Locations of each stop (lat, long and stop_id)
stop_times.txt: Arrival times at each stop for each trip (stop_id and trip_id)
trips.txt: Tells you which trips belong to which route and service pattern (trip_id and service_id and route_id)
calendar.txt and calendar_dates.txt which service patterns are running on which days.

#Los Angeles

## Read your GTFS data into R:

```{r}
la_stops_path <- "../02-data/LA_Metro_4OCT2019/stops.txt"

la_stops <- read_csv(la_stops_path) %>%
  select(stop_id, stop_lon, stop_lat)

la_times_path <- "../02-data/LA_Metro_4OCT2019/stop_times.txt"

la_times <- read_csv(la_times_path)

la_routes_path <- "../02-data/LA_Metro_4OCT2019/routes.txt"

la_routes <- read_csv(la_routes_path)

la_times <- la_times %>%
  separate(stop_headsign, into=c("route_short_name","route_desc"), sep=" - ", extra = "merge")

la_stop_sums <- la_times %>%
  group_by(stop_id, route_short_name) %>%
  tally()

la_stop_sums_freq <- la_stop_sums %>%
  left_join(la_stops) %>%
  slice_max(n)
  
la_stops_sums_freq_sf = st_as_sf(la_stop_sums_freq, coords = c("stop_lon", "stop_lat"), 
                 crs = 4326, agr = "constant")  
```

## Define study area

The study area is defined as the set of Los Angeles County zip code tabulation areas within 500 meters of an LA Metro transit stop. Five zip codes were excluded in order to more closely align a census-block-based study area with a zip-code-based study area. 

```{r}
la_co = counties(state = "CA", year = 2010) %>%
  filter(NAME10 == "Los Angeles")

la_zctas <- get_decennial(geography = "zcta",
                           state = "CA",
                           variables = demographics,
                           year = 2010,
                           output = "wide",
                          geometry = TRUE) %>%
  st_filter(la_co) %>%
  mutate(area = st_area(.))

la_nstops <- lengths(st_is_within_distance(la_zctas, la_stops_sums_freq_sf, dist = 500))

la_zctas <- la_zctas %>%
  filter(la_nstops > 0) %>%
  filter(NAME != "ZCTA5 90265, California" & # for boundaries to match
           NAME != "ZCTA5 91361, California" & # for boundaries to match
           NAME != "ZCTA5 91360, California" & # for boundaries to match
           NAME != "ZCTA5 91362, California" & # for boundaries to match
           NAME != "ZCTA5 91307, California" & # for boundaries to match
           NAME != "ZCTA5 90263, California" & # for a contiguous study area
           NAME != "ZCTA5 92833, California" & # Orange county
           NAME != "ZCTA5 90621, California" & # Orange county
           NAME != "ZCTA5 90620, California") %>% # Orange county
  st_transform("WGS84")

#st_is_within_distance(x, y = x, dist, sparse = TRUE, ...)

la_zctas <- la_zctas %>%
  #mutate(non_white_pop = blk_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  mutate(la_tot_pop = blk_pop + white_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop)

la_zctas %>%
  select(NAME) %>%
  st_drop_geometry() %>%
  write_csv(file = "../02-data/la-area-zips.csv")
```

## Map study area

Here is an interactive map of the zip code tabulation areas in the study area.

```{r}
la_map <- leaflet(la_zctas) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = TRUE),
      weight = 1,
      fillColor = "green",
      color = "green",
      label = ~ NAME)

here("03-maps",
     "zips.html") %>%  
  saveWidget(widget = la_map, .)
```

# Load census blocks

Get a list of all the census blocks in the study area and assign each block to the 
closest route.

```{r}
la_blocks <- get_decennial(geography="block",
                state = "CA", 
                county = "Los Angeles", 
                variables = demographics,
                year = 2010,
                geometry = TRUE) %>%
  st_transform(st_crs(la_zctas)) %>%
  mutate(area = st_area(.))

blocks_in_la_zctas <- st_centroid(la_blocks) %>%
  st_filter(la_zctas) %>%
  select(GEOID) %>%
  st_drop_geometry()
  
la_blocks_filtered <- la_blocks %>%
  filter(GEOID %in% blocks_in_la_zctas$GEOID) %>%
  #right_join(block_in_zctas) %>%
  st_join(la_stops_sums_freq_sf, join = st_nearest_feature) %>%
  select(GEOID, route_short_name, variable, value) %>%
  pivot_wider(names_from = "variable",
              values_from = "value")

la_routes_summed <- la_blocks_filtered %>%
  #mutate(non_white_pop = blk_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  mutate(la_tot_pop = blk_pop + white_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  st_drop_geometry() %>%
  group_by(route_short_name) %>% 
  summarise(across(.cols = -GEOID, sum))
```

# Create route zone boundaries

The summarize function takes a very long time. I'm saving the result so you can just read it from file and skip this chunk.

```{r, eval=FALSE}
la_route_zones <- la_blocks_filtered %>%
  group_by(route_short_name) %>%
  summarise()

here("02-data",
     "la-route-zones.geojson") %>%  
  st_write(obj = la_route_zones, .)
```

## Map the boundaries

```{r}
la_route_zones <- here("02-data",
                    "la-route-zones.geojson") %>%
  st_read()
la_map <- leaflet(la_route_zones) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = FALSE),
      weight = 1,
      color = "green",
      fillColor = "green",
      label = la_route_zones$route_short_name) 

here("03-maps",
     "la-route-zones.html") %>%  
  saveWidget(widget = la_map, .)
```

# Save the list of blocks

```{r}
la_blocks_filtered %>%
  st_drop_geometry() %>%
  write_csv(here("02-data",
                 "la-blocks-by-route.csv"))
```

#Boston

## 1.

```{r}
mbta_stops_path <- "../02-data/BOSTON_MBTA_4OCT2019/stops.txt"

mbta_stops <- read_csv(mbta_stops_path) %>%
  select(stop_id, stop_lon, stop_lat)

mbta_times_path <- "../02-data/BOSTON_MBTA_4OCT2019/stop_times.txt"

mbta_times <- read_csv(mbta_times_path)

mbta_trips <- read_csv("../02-data/BOSTON_MBTA_4OCT2019/trips.txt") %>%
  select(trip_id, route_id)

mbta_routes_path <- "../02-data/BOSTON_MBTA_4OCT2019/routes.txt"

mbta_routes <- read_csv(mbta_routes_path) %>%
  filter(route_desc %in% c("Key Bus", "Local Bus"))

mbta_stops_sums <- mbta_trips %>%
  left_join(mbta_times) %>%
  group_by(stop_id, route_id) %>%
  tally() %>%
  left_join(mbta_stops)

mbta_stops_sums_sf = st_as_sf(mbta_stops_sums, coords = c("stop_lon", "stop_lat"), 
                 crs = 4326, agr = "constant")  
```

## 2.
# MBTA Counties
```{r mbta_counties}
mbta_co = counties(state = "MA", year = 2010) 

mbta_co <- mbta_co %>%
  filter(mbta_co$NAME10 %in% c("Barnstable","Bristol", "Worcester", "Suffolk", "Middlesex", "Norfolk", "Plymouth", "Essex"))

mbta_zctas <- get_decennial(geography = "zcta",
                           state = "MA",
                           variables = demographics,
                           year = 2010,
                           output = "wide",
                          geometry = TRUE) %>%
  st_filter(mbta_co) %>%
  mutate(area = st_area(.))

mbta_nstops <- lengths(st_is_within_distance(mbta_zctas, mbta_stops_sums_sf, dist = 500))

mbta_zctas <- mbta_zctas %>%
 filter(mbta_nstops > 0) %>%
  #filter(NAME != "ZCTA5 90265, California" & # for boundaries to match
          #NAME != "ZCTA5 91361, California" & # for boundaries to match
          #NAME != "ZCTA5 91360, California" & # for boundaries to match
          #NAME != "ZCTA5 91362, California" & # for boundaries to match
          #NAME != "ZCTA5 91307, California" & # for boundaries to match
          #NAME != "ZCTA5 90263, California" & # for a contiguous study area
          #NAME != "ZCTA5 92833, California" & # Orange county
          #NAME != "ZCTA5 90621, California" & # Orange county
          #NAME != "ZCTA5 90620, California") %>% # Orange county
  st_transform("WGS84")

#st_is_within_distance(x, y = x, dist, sparse = TRUE)

mbta_zctas <- mbta_zctas %>%
  #mutate(non_white_pop = blk_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  mutate(mbta_tot_pop = blk_pop + white_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop)

mbta_zctas %>%
  select(NAME) %>%
  st_drop_geometry() %>%
  write_csv(file = "../02-data/mbta-area-zips.csv")
```

## 3. Use the dplyr package to join the stops_tidy and stop_times_tidy data frames on the stop_id column and group the data by stop_id and route_id:

```{r}
mbta_map <- leaflet(mbta_zctas) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = TRUE),
      weight = 1,
      fillColor = "green",
      color = "green",
      label = ~ NAME)

here("03-maps",
     "mbta-zips.html") %>%  
  saveWidget(widget = mbta_map, .)
```

## 4. Use the summarize() function to count the number of trips per route at each stop:

```{r}
mbta_blocks <- get_decennial(geography="block",
                state = "MA", 
                county = c("Barnstable","Bristol", "Worcester", "Suffolk", "Middlesex", "Norfolk", "Plymouth", "Essex"),
                variables = demographics,
                year = 2010,
                geometry = TRUE) %>%
  st_transform(st_crs(mbta_zctas)) %>%
  mutate(area = st_area(.))

mbta_blocks_in_zctas <- st_centroid(mbta_blocks) %>%
  st_filter(mbta_zctas) %>%
  select(GEOID) %>%
  st_drop_geometry()
  
mbta_blocks_filtered <- mbta_blocks %>%
  filter(GEOID %in% mbta_blocks_in_zctas$GEOID) %>%
  #right_join(block_in_zctas) %>% 
  st_join(mbta_stops_sums_sf, join = st_nearest_feature) %>%
  select(GEOID, variable, value, route_id) %>%
  pivot_wider(names_from = "variable",
              values_from = "value")

mbta_routes_summed <- mbta_blocks_filtered %>%
  #mutate(non_white_pop = blk_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  mutate(mbta_tot_pop = blk_pop + white_pop + hispanic_pop + asian_pop + nat_am_pop + nhpi_pop) %>%
  st_drop_geometry() %>%
  group_by(route_id) %>% 
  summarise(across(.cols = -GEOID, sum))
```

## 5. Use the dplyr function arrange() to sort the data frame by the n column in descending order:

```{r}
mbta_route_zones <- mbta_blocks_filtered %>%
  group_by(route_id) %>%
  summarise()

here("02-data",
     "mbta-route-zones.geojson") %>%  
  st_write(obj = mbta_route_zones, .)
```

## 6. Use the dplyr function slice() to select the first row of each group (i.e., the route with the highest number of trips at each stop):

```{r}
mbta_route_zones <- here("02-data",
                    "mbta-route-zones.geojson") %>%
  st_read()
mbta_map <- leaflet(mbta_route_zones) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = FALSE),
      weight = 1,
      color = "green",
      fillColor = "green",
      label = mbta_route_zones$route_id) 

here("03-maps",
     "mbta-route-zones.html") %>%  
  saveWidget(widget = mbta_map, .)
```

```{r}
mbta_blocks %>%
  st_drop_geometry() %>%
  write_csv(here("02-data",
                 "mbta-blocks-by-route.csv"))
```

#Chicago


## 1. Read your GTFS data into R using the here() function:

```{r}
cta_stops_path <- "../02-data/CHICAGO_CTA_4OCT2019/stops.txt"

cta_stops <- read_csv(cta_stops_path) %>%
  select(stop_id, stop_lon, stop_lat)

cta_times_path <- "../02-data/CHICAGO_CTA_4OCT2019/stop_times.txt"

cta_times <- read_csv(cta_times_path) %>%
  mutate(trip_id = as.character(trip_id))

cta_routes_path <- "../02-data/CHICAGO_CTA_4OCT2019/routes.txt"

cta_routes <- read_csv(cta_routes_path) 
  
cta_stop_times_path <- "../02-data/CHICAGO_CTA_4OCT2019/stop_times.txt"

cta_stop_times <- read_csv(cta_stop_times_path)

cta_trips_path <- "../02-data/CHICAGO_CTA_4OCT2019/trips.txt"

cta_trips <- read_csv(cta_trips_path, col_types = cols(.default = col_character())) %>%
  select(trip_id, route_id)

cta_stops_sums <- cta_trips %>%
  left_join(cta_times) %>%
  group_by(stop_id, route_id) %>%
  tally() %>%
  left_join(cta_stops)

cta_stop_sums_freq <- cta_stops_sums %>%
  group_by(stop_id) %>%
  slice_max(n) %>%
  left_join(cta_stops)
  
cta_stops_sums_freq_sf = st_as_sf(cta_stop_sums_freq, coords = c("stop_lon", "stop_lat"), 
                 crs = 4326, agr = "constant")  
  
```

## 2.

```{r}
cta_co = counties(state = "IL", year = 2010)

cta_co <- cta_co %>%
  filter(cta_co$NAME10 %in% c("Lake", "Cook", "Kane", "DuPage", "Will")) 

cta_zctas <- get_decennial(geography = "zcta",
                           state = "IL",
                           variables = demographics,
                           year = 2010,
                           output = "wide",
                          geometry = TRUE) %>%
  st_filter(cta_co) %>%
  mutate(area = st_area(.))

cta_nstops <- lengths(st_is_within_distance(cta_zctas, cta_stops_sums_freq_sf, dist = 500))

cta_zctas <- cta_zctas %>%
 filter(cta_nstops > 0) %>%
  #filter(NAME != "ZCTA5 90265, California" & # for boundaries to match
          #NAME != "ZCTA5 91361, California" & # for boundaries to match
          #NAME != "ZCTA5 91360, California" & # for boundaries to match
          #NAME != "ZCTA5 91362, California" & # for boundaries to match
          #NAME != "ZCTA5 91307, California" & # for boundaries to match
          #NAME != "ZCTA5 90263, California" & # for a contiguous study area
          #NAME != "ZCTA5 92833, California" & # Orange county
          #NAME != "ZCTA5 90621, California" & # Orange county
          #NAME != "ZCTA5 90620, California") %>% # Orange county
  st_transform("WGS84")

#st_is_within_distance(x, y = x, dist, sparse = TRUE)

cta_zctas %>%
  select(NAME) %>%
  st_drop_geometry() %>%
  write_csv(file = "../02-data/cta-area-zips.csv")
```

## 3. Use the dplyr package to join the stops_tidy and stop_times_tidy data frames on the stop_id column and group the data by stop_id and route_id:

```{r}
cta_map <- leaflet(cta_zctas) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = TRUE),
      weight = 1,
      fillColor = "green",
      color = "green",
      label = ~ NAME)

here("03-maps",
     "cta-zips.html") %>%  
  saveWidget(widget = cta_map, .)
```

## 4. Use the summarize() function to count the number of trips per route at each stop:

```{r}
cta_blocks <- get_decennial(geography="block",
                state = "IL", 
                county = c("Lake", "Cook", "Kane", "DuPage", "Will County"),
                variables = demographics,
                year = 2010,
                geometry = TRUE) %>%
  st_transform(st_crs(cta_zctas)) %>%
  mutate(area = st_area(.))

cta_blocks_in_zctas <- st_centroid(cta_blocks) %>%
  st_filter(cta_zctas) %>%
  select(GEOID) %>%
  st_drop_geometry()
  
cta_blocks_filtered <- cta_blocks %>%
  filter(GEOID %in% cta_blocks_in_zctas$GEOID) %>%
  #right_join(block_in_zctas) %>% 
  st_join(cta_stops_sums_freq_sf, join = st_nearest_feature) %>%
  select(GEOID, route_id, variable, value) %>%
  pivot_wider(names_from = "variable",
              values_from = "value")
```

## 5. Use the dplyr function arrange() to sort the data frame by the n column in descending order:

```{r}
cta_route_zones <- cta_blocks_filtered %>%
  group_by(route_id) %>%
  summarise()

here("02-data",
     "cta-route-zones.geojson") %>%  
  st_write(obj = cta_route_zones, .)
```

## 6. Use the dplyr function slice() to select the first row of each group (i.e., the route with the highest number of trips at each stop):

```{r}
cta_route_zones <- here("02-data",
                    "cta-route-zones.geojson") %>%
  st_read()
cta_map <- leaflet(cta_route_zones) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = FALSE),
      weight = 1,
      color = "green",
      fillColor = "green",
      label = cta_route_zones$route_id) 

here("03-maps",
     "cta-route-zones.html") %>%  
  saveWidget(widget = cta_map, .)
```

```{r}
cta_blocks %>%
  st_drop_geometry() %>%
  write_csv(here("02-data",
                 "cta-blocks-by-route.csv"))
```

# Dallas


## 1. Read your GTFS data into R using the here() function:

```{r}
dart_stops_path <- "../02-data/DALLAS_DART_4OCT2019/stops.txt"

dart_stops <- read_csv(dart_stops_path) %>%
  select(stop_id, stop_lon, stop_lat)

dart_times_path <- "../02-data/DALLAS_DART_4OCT2019/stop_times.txt"

dart_times <- read_csv(dart_times_path) %>%
    mutate(trip_id = as.character(trip_id))

dart_routes_path <- "../02-data/DALLAS_DART_4OCT2019/routes.txt"

dart_routes <- read_csv(dart_routes_path)

dart_trips_path <- "../02-data/DALLAS_DART_4OCT2019/trips.txt"

dart_trips <- read_csv(dart_trips_path, col_types = cols(.default = col_character())) %>%
  select(trip_id, route_id)

dart_stops_sums <- dart_trips %>%
  left_join(dart_times) %>%
  group_by(stop_id, route_id) %>%
  tally() %>%
  left_join(dart_stops)

dart_stop_sums_freq <- dart_stops_sums %>%
  group_by(stop_id) %>%
  slice_max(n)
  
dart_stops_sums_freq_sf = st_as_sf(cta_stop_sums_freq, coords = c("stop_lon", "stop_lat"), 
                 crs = 4326, agr = "constant")

```

## 2.
# DART Counties ------
```{r mbta_counties}
dart_co = counties(state = "TX", year = 2010)

dart_co <- dart_co %>%
  filter(dart_co$NAME10 %in% c("Dallas", "Tarrant", "Collin", "Denton", "Ellis", "Hunt", "Rockwall", "Kaufman"))

dart_zctas <- get_decennial(geography = "zcta",
                           state = "TX",
                           variables = demographics,
                           year = 2010,
                           output = "wide",
                          geometry = TRUE) %>%
  st_filter(dart_co) %>%
  mutate(area = st_area(.))

dart_nstops <- lengths(st_is_within_distance(dar_zctas, dart_stops_sums_sf, dist = 500))

dart_zctas <- dart_zctas %>%
 filter(dart_nstops > 0) %>%
  #filter(NAME != "ZCTA5 90265, California" & # for boundaries to match
          #NAME != "ZCTA5 91361, California" & # for boundaries to match
          #NAME != "ZCTA5 91360, California" & # for boundaries to match
          #NAME != "ZCTA5 91362, California" & # for boundaries to match
          #NAME != "ZCTA5 91307, California" & # for boundaries to match
          #NAME != "ZCTA5 90263, California" & # for a contiguous study area
          #NAME != "ZCTA5 92833, California" & # Orange county
          #NAME != "ZCTA5 90621, California" & # Orange county
          #NAME != "ZCTA5 90620, California") %>% # Orange county
  st_transform("WGS84")

#st_is_within_distance(x, y = x, dist, sparse = TRUE)

dart_zctas %>%
  select(NAME) %>%
  st_drop_geometry() %>%
  write_csv(file = "../02-data/study-area-zips.csv")
```

## 3. Use the dplyr package to join the stops_tidy and stop_times_tidy data frames on the stop_id column and group the data by stop_id and route_id:

```{r}
dart_map <- leaflet(dart_zctas) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = TRUE),
      weight = 1,
      fillColor = "green",
      color = "green",
      label = ~ NAME)

here("03-maps",
     "dart-zips.html") %>%  
  saveWidget(widget = dart_map, .)
```

## 4. Use the summarize() function to count the number of trips per route at each stop:

```{r}
dart_blocks <- get_decennial(geography="block",
                state = "TX", 
                county = x,
                variables = demographics,
                year = 2010,
                geometry = TRUE) %>%
  st_transform(st_crs(dart_zctas))

dart_blocks_in_zctas <- st_centroid(dart_blocks) %>%
  st_filter(dart_zctas) %>%
  select(GEOID) %>%
  st_drop_geometry()
  
dart_blocks_filtered <- dart_blocks %>%
  filter(GEOID %in% dart_blocks_in_zctas$GEOID) %>%
  #right_join(block_in_zctas) %>% 
  st_join(dart_stops_sums_sf, join = st_nearest_feature) %>%
  select(GEOID, route_short_name, variable, value) %>%
  pivot_wider(names_from = "variable",
              values_from = "value")
```

## 5. Use the dplyr function arrange() to sort the data frame by the n column in descending order:

```{r}
dart_route_zones <- dart_blocks_filtered %>%
  group_by(route_short_name) %>%
  summarise()

here("02-data",
     "dart-route-zones.geojson") %>%  
  st_write(obj = dart_route_zones, .)
```

## 6. Use the dplyr function slice() to select the first row of each group (i.e., the route with the highest number of trips at each stop):

```{r}
dart_route_zones <- here("02-data",
                    "dart-route-zones.geojson") %>%
  st_read()
dart_map <- leaflet(dart_route_zones) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(highlightOptions = highlightOptions(
      weight = 1,
      fillOpacity = 1,
      bringToFront = FALSE),
      weight = 1,
      color = "green",
      fillColor = "green",
      label = dart_route_zones$route_short_name) 

here("03-maps",
     "dart-route-zones.html") %>%  
  saveWidget(widget = dart_map, .)
```

```{r}
dart_blocks %>%
  st_drop_geometry() %>%
  write_csv(here("02-data",
                 "dart-blocks-by-route.csv"))
```

# Measures of Segregation

pi = the ratio of xi to ti (proportion of area i's population that is minority)
xi = the minority population of area i
ti = the total population of area i
yi= the majority population (non-Hispanic Whites in this report) of area i
X = the sum of all xi (the total minority population)
T = the sum of all ti (the total population)
P = the ratio of X to T (proportion of the metropolitan area's population that is minority)

# The "Definitions" included in the paper are super useful, but they
# break a bunch of math conventions that make using their variable
# names confusing

# So I'm going to write this the way a professional coder would expect
# it to be written: only the input that's needed gets passed in to the
# function, which in this case is the list of minority populations in
# areas (this would be the list of x_i's in the paper)

## IMPORTANT: for relative_concentation_index() to work, sort the 
# min_pops_by_area (x_i) and tot_pops_by_area (t_i) by land_area_by_area (a_i)
# from least (a_1) to greatest (a_1) => that's the only way the ranking works at
# present

# Index of Dissimilarity

```{r}
# dissimilarity // x_i = min_pops_by_area (list); t_i = tot_pops_by_area (list)
dissimilarity_index <- function(min_pops_by_area, tot_pops_by_area) {
  proportions_by_area <- min_pops_by_area / tot_pops_by_area # pi
  proportions_by_area <- na.omit(proportions_by_area) # drop division errors / NAs
  tot_min_pop <- sum(min_pops_by_area) # X
  overall_pop <- sum(tot_pops_by_area) # T
  overall_prop <- tot_min_pop / overall_pop # P

  numerator <- sum(tot_min_pop * abs(proportions_by_area - overall_prop))
  denominator <- 2*overall_pop*overall_prop*(1 - overall_prop)
  dissimilarity <- numerator / denominator
  return(dissimilarity)
}
```

## LA Hispanic Population

```{r}
dissimilarity_index(min_pops_by_area = la_zctas$hispanic_pop, tot_pops_by_area = la_zctas$la_tot_pop)
dissimilarity_index(min_pops_by_area = la_routes_summed$hispanic_pop, tot_pops_by_area = la_routes_summed$la_tot_pop)
```

## LA Black Population

```{r}
dissimilarity_index(min_pops_by_area = la_zctas$blk_pop, tot_pops_by_area = la_zctas$la_tot_pop)
dissimilarity_index(min_pops_by_area = la_routes_summed$blk_pop, tot_pops_by_area = la_routes_summed$la_tot_pop)
```

## LA Asian Population

```{r}
dissimilarity_index(min_pops_by_area = la_zctas$asian_pop, tot_pops_by_area = la_zctas$la_tot_pop)
dissimilarity_index(min_pops_by_area = la_routes_summed$asian_pop, tot_pops_by_area = la_routes_summed$la_tot_pop)
```

## Boston Hispanic Population

```{r}
dissimilarity_index(min_pops_by_area = mbta_zctas$hispanic_pop, tot_pops_by_area = mbta_zctas$mbta_tot_pop)
dissimilarity_index(min_pops_by_area = mbta_routes_summed$hispanic_pop, tot_pops_by_area = mbta_routes_summed$mbta_tot_pop)
```

## Boston Black Population

```{r}
dissimilarity_index(min_pops_by_area = mbta_zctas$blk_pop, tot_pops_by_area = mbta_zctas$mbta_tot_pop)
dissimilarity_index(min_pops_by_area = mbta_routes_summed$blk_pop, tot_pops_by_area = mbta_routes_summed$mbta_tot_pop)
```

## Boston Asian Population

```{r}
dissimilarity_index(min_pops_by_area = mbta_zctas$asian_pop, tot_pops_by_area = mbta_zctas$mbta_tot_pop)
dissimilarity_index(min_pops_by_area = mbta_routes_summed$asian_pop, tot_pops_by_area = mbta_routes_summed$mbta_tot_pop)
```

# Index of isolation

```{r}
# isolation // x_i = min_pops_by_area (list); t_i = tot_pops_by_area (list)
isolation_index <- function(min_pops_by_area, tot_pops_by_area) {
  tot_min_pop <- sum(min_pops_by_area) # X
  fraction_of_min_pop_by_area = min_pops_by_area / tot_min_pop # x_i / X
  min_to_total_ratios = min_pops_by_area / tot_pops_by_area # x_i / t_i
  return(sum(fraction_of_min_pop_by_area * min_to_total_ratios, na.rm = TRUE))
}
```

## LA Hispanic Population

```{r}
isolation_index(la_zctas$hispanic_pop, la_zctas$la_tot_pop)
isolation_index(la_routes_summed$hispanic_pop, la_routes_summed$la_tot_pop)
```

## LA Black Population

```{r}
isolation_index(la_zctas$blk_pop, la_zctas$la_tot_pop)
isolation_index(la_routes_summed$blk_pop, la_routes_summed$la_tot_pop)
```

## LA Asian Population

```{r}
isolation_index(la_zctas$asian_pop, la_zctas$la_tot_pop)
isolation_index(la_routes_summed$asian_pop, la_routes_summed$la_tot_pop)
```

## Boston Hispoanic Population

```{r}
isolation_index(mbta_zctas$hispanic_pop, mbta_zctas$mbta_tot_pop)
isolation_index(mbta_routes_summed$hispanic_pop, mbta_routes_summed$mbta_tot_pop)
```

## Boston Black Population

```{r}
isolation_index(mbta_zctas$blk_pop, mbta_zctas$mbta_tot_pop)
isolation_index(mbta_routes_summed$blk_pop, mbta_routes_summed$mbta_tot_pop)
```

## Boston Asian Population

```{r}
isolation_index(mbta_zctas$asian_pop, mbta_zctas$mbta_tot_pop)
isolation_index(mbta_routes_summed$asian_pop, mbta_routes_summed$mbta_tot_pop)
```

# Index of Relative Concentration

```{r}
# relative concentration // x_i = min_pops_by_area (list); t_i = tot_pops_by_area (list); y_i = maj_pops_by_area; a_i = land_area_by_area
relative_concentation_index <- function(min_pops_by_area, tot_pops_by_area, maj_pops_by_area, land_area_by_area) {
  tot_min_pop <- sum(min_pops_by_area) # X
  tot_maj_pop <- sum(maj_pops_by_area) # Y
  
  top_fraction <- sum((min_pops_by_area * land_area_by_area) / tot_min_pop) / sum((maj_pops_by_area * land_area_by_area) / tot_maj_pop)
  numerator <- top_fraction - 1
  
  n <- length(min_pops_by_area)
  # n1 = rank of area where the sum of all ti from area 1 (smallest in size) 
  # up to area n1 is equal to X
  n1 <- 0
  running_total_sum_for_n1 <- 0
  for (pop in tot_pops_by_area) {
    if (running_total_sum_for_n1 + pop > tot_min_pop) {
      break
    } else {
      running_total_sum_for_n1 <- running_total_sum_for_n1 + pop
      n1 <- n1 + 1
    }
  }

    # n2 = rank of area where the sum of all ti from area n (largest in size) 
  # down to area n2 is equal to X
  n2_offset <- 0
  running_total_sum_for_n2 <- 0
  for (pop in tot_min_pop) {
    if (running_total_sum_for_n2 + pop > tot_min_pop) {
      break
    } else {
      running_total_sum_for_n2 <- running_total_sum_for_n2 + pop
      n2_offset <- n2_offset + 1
    }
  }
  
  n2 <- n - n2_offset
  # T_1 = the sum of all ti in area 1 up to area n1
  T_1 <- tot_pops_by_area[1:n1]
  # T_2 = the sum of all ti in area n2 up to area n
  T_2 <- tot_pops_by_area[n2:n]
  bottom_fraction <- sum((tot_pops_by_area[1:n1] * land_area_by_area[1:n1]) / T_1) / sum((tot_pops_by_area[n2:n] * land_area_by_area[n2:n]) / T_2)
  denominator <- bottom_fraction - 1
  
  return(numerator / denominator)
}

relative_concentation_index(c(5, 10, 15, 200), c(60, 60, 60, 60), c(50, 30, 30, 300), c(8000, 600, 400, 200))
relative_concentation_index(c(500, 100, 150, 200), c(60, 60, 60, 60), c(600, 300, 300, 300), c(8000, 600, 400, 200))
```


## Calculate accessibility

```{r}
# Load the required libraries
library(igraph)
library(httr)

# Define the API key
api_key <- "b8064d02a4943b2c25e30fca13e816fc"

# Define the base URL for the API
base_url <- "https://api.metro.net"

# Define the endpoint for the bus routes
routes_endpoint <- "/api/v3/routes"

# Make a request to the API to retrieve the bus routes
routes_response <- GET(paste0(base_url, routes_endpoint), query = list(api_key = api_key))

# Parse the JSON response into a data frame
routes_data <- content(routes_response, as = "parsed")$items

# Filter the routes data to only include bus routes
bus_routes_data <- routes_data[routes_data$mode == "bus",]

# Define the endpoint for the stops
stops_endpoint <- "/api/v3/stops"

# Make a request to the API to retrieve the stops
stops_response <- GET(paste0(base_url, stops_endpoint), query = list(api_key = api_key))

# Parse the JSON response into a data frame
stops_data <- content(stops_response, as = "parsed")$items

# Create a graph object representing the LA Metro bus network
bus_network_graph <- graph_empty(n = nrow(stops_data))

# Add edges between the stops to represent the bus routes
for (i in 1:nrow(bus_routes_data)) {
  route_id <- bus_routes_data[i, "id"]
  route_endpoint <- paste0("/api/v3/routes/", route_id, "/sequence")
  route_response <- GET(paste0(base_url, route_endpoint), query = list(api_key = api_key))
  route_data <- content(route_response, as = "parsed")$items
  
  for (j in 1:(length(route_data) - 1)) {
    stop1 <- route_data[j]
    stop2 <- route_data[j + 1]
    edge_id <- c(stop1$id, stop2$id)
    edge_weight <- distHaversine(c(stop1$latitude, stop1$longitude), c(stop2$latitude, stop2$longitude))
    E(bus_network_graph) <- c(E(bus_network_graph), edge_id)
    set.edge.attribute(bus_network_graph, name = "weight", index = E(bus_network_graph)[length(E(bus_network_graph))], value = edge_weight)
  }
}

# Define a function to calculate the accessibility measure for a given origin and destination
calculate_accessibility <- function(origin, destination) {
  shortest_path <- shortest.paths(bus_network_graph, origin, destination, weights = E(bus_network_graph)$weight, output = "epath")
  accessibility <- sum(E(bus_network_graph)$weight[shortest_path[[1]]])
  return(accessibility)
}
```

```{r}
library(ggplot2)

# Create a data frame to store the accessibility values
accessibility_data <- data.frame(origin = character(), destination = character(), accessibility = numeric())

# Loop over all combinations of origin and destination stops
for (i in 1:nrow(stops_data)) {
  for (j in 1:nrow(stops_data)) {
    # Calculate the accessibility between the origin and destination
    accessibility <- calculate_accessibility(i, j)
    # Add the origin, destination, and accessibility values to the data frame
    accessibility_data <- rbind(accessibility_data, data.frame(origin = stops_data$stop_id[i], destination = stops_data$stop_id[j], accessibility = accessibility))
  }
}

# Plot the accessibility data as a heat map
ggplot(data = accessibility_data, aes(x = origin, y = destination, fill = accessibility)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  xlab("Origin") +
  ylab("Destination") +
  ggtitle("Accessibility between Bus Stops in LA Metro")
```